---
title: "MLAssignment3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Install package if not already installed
```{r}
#install.packages("caret")
#install.packages("lattice")
#install.packages("ggplot2")
#install.packages("e1071")
#install.packages("klaR")
#install.packages("psych")
#install.packages("rpivotTable")
```


#load all the required libraries
```{r, results='hide'}
library(caret)
library(readr)
library(gmodels)
library(ISLR)
library(e1071) 
library(dplyr)
library(pROC)
library(rpivotTable)
```


# Import the FlightDelays Dataset
```{r}
FlightData <- read.csv("FlightDelays.csv")
summary(FlightData)
```


# Record Week columns as fators
```{r}
FlightData$DAY_WEEK <- factor(FlightData$DAY_WEEK)
levels(FlightData$DAY_WEEK)
```


# Create bins for Scheduled Departure Time column and record it as factors
```{r}
FlightData$CRS_DEP_TIME <- floor(FlightData$CRS_DEP_TIME/100)
FlightData$CRS_DEP_TIME <- factor(FlightData$CRS_DEP_TIME)
levels(FlightData$CRS_DEP_TIME)
```


# Label the outcome variable in two classes (1 = Delayed, 0 = Ontime)
```{r}
FlightData$Flight.Status <- factor(FlightData$Flight.Status, levels = c("ontime","delayed"), labels = c(0,1))
levels(FlightData$Flight.Status)
```


# Partition the data into training (60%) and validation (40%) sets
```{r}
set.seed(123)
Train_Index=createDataPartition(FlightData$CRS_DEP_TIME, p=0.60, list=FALSE)
Train_Data = FlightData[Train_Index,] 
Validation_Data = FlightData[-Train_Index,] 
```


# Now, run the Naive Bayes model on the training dataset
```{r}
nb_model <- naiveBayes(Train_Data$Flight.Status~CARRIER+DEST+ORIGIN+DAY_WEEK+CRS_DEP_TIME, data = Train_Data)
nb_model
```

***
# The output from the Naive Bayes model predicts that the probability of Delayed flight is 0.2019593 and the probability for On time flights is 0.7980407 
***


# Output Counts table and proportion table for flights that are delayed and ontime from Origin
```{r}
#Counts table 
rpivotTable(FlightData, rows = "Flight.Status", cols = "ORIGIN", width = "100%", height = "400px")

#Proportion Table
prop.table(table(FlightData$Flight.Status, FlightData$ORIGIN),margin = 1)
```


# Output Counts table and proportion table for ontime and delayed flights at Destination
```{r}
#Counts table
rpivotTable(FlightData, rows = "Flight.Status", cols = "DEST", width = "100%", height = "400px")

#Proportion Table
prop.table(table(FlightData$Flight.Status, FlightData$DEST),margin = 1)
```


# Confusion matrix for the validation data
```{r}
#Make predictions and return probability 
PredictData <-predict(nb_model,Validation_Data)

#show the first few values
head(PredictData)

#Confusion Matrix
confusionMatrix(PredictData, Validation_Data$Flight.Status)
```

***
# The Confusion matrix for the validation data set shows the Accuracy of 0.8089
***


# Output the ROC for the validation data
```{r}
#Make predictions and return probability of each class
PredictData <-predict(nb_model,Validation_Data, type = "raw")

#show the first few values
head(PredictData)

#ROC Curve for validation Data set
roc(Validation_Data$Flight.Status, PredictData[,2])
plot.roc(Validation_Data$Flight.Status,PredictData[,2])
```

***
# The Area Under the Curve is 0.661
***